---
title: "Lung_Cancer_Analysis"
output: html_document
date: "2025-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Set Up

## Install and Load Packages
```{r}

library(tidyverse) # load tidyverse package to make graphs and format the data in tables
library(car) # load car package for vifs to test for multicollinearity in multiple linear regressions

```

## Load Data for the Lung Cancer Analysis from the Main Project Folder
```{r}

# Load the CSV files for Lung Cancer Registrations vs Deprivation
lungcancer_registrations_deprivation <- read.csv("Lung_cancer_registrations_vs_deprivation.csv")
lungcancer_registrations_deprivation

# Load the CSV files for Lung Cancer Mortality Rate vs Deprivation
lungcancer_mortality_deprivation <- read.csv("Lung_cancer_mortality_vs_deprivation.csv")
lungcancer_mortality_deprivation 

# Load the CSV files for Lung Cancer Mortality Rate vs Lung Cancer Registrations
lungcancer_registrations_mortality <- read.csv("lungcancer_registrations_mortality.csv")
lungcancer_registrations_mortality

# Load the CSV files for Smoking Prevalence vs Deprivation
smoking_deprivation <- read.csv("Smoking_deprivation.csv")
smoking_deprivation

# Load the CSV files for Smoking Prevalence vs Lung Cancer Registrations
smoking_lung_cancer_registrations <- read.csv("smoking_vs_registrations.csv")
smoking_lung_cancer_registrations

# Load the CSV files for Smoking Prevalence vs Lung Cancer Registrations
smoking_lung_cancer_mortality <- read.csv("smoking_mortality.csv")
smoking_lung_cancer_mortality

# Load CSV Files for Smoking Prevalence, Deprivation and Lung Cancer Registrations

smoking_deprivation_lung_cancer_registrations <- read.csv("lungcancer_deprivation_registrations_smoking.csv")
smoking_deprivation_lung_cancer_registrations


# Load CSV Files for Smoking Prevalence, Deprivation and  Lung Cancer Registrations

smoking_deprivation_lung_cancer_mortality <- read.csv("smoking_deprivation_mortality.csv")
smoking_deprivation_lung_cancer_mortality


```


## Data Processing
```{r}

# Function to rename the columns in a dataset  
rename_columns <- function(dataset, new_colnames) {  
   
   # Check if the number of new column names matches the number of columns in the dataset  
   if (ncol(dataset) != length(new_colnames)) {  
     stop("Error: The number of columns in the dataset does not match the provided column names.")  
     # stop() halts execution and returns an error message if the column counts do not match  
   }  
   
   # Assign the new column names to the dataset  
   colnames(dataset) <- new_colnames  
   
   # Return the modified dataset with the updated column names  
   return(dataset)  
}  

# Rename Columns for Each Dataset
lungcancer_registrations_deprivation <- rename_columns(
  lungcancer_registrations_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)", "Lung Cancer Registrations per 100,000 (2017-2019)")
)

lungcancer_mortality_deprivation <- rename_columns(
  lungcancer_mortality_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)", "Lung Cancer Mortality Rate per 100,000 (2021-2023)")
)

lungcancer_registrations_mortality <- rename_columns(
  lungcancer_registrations_mortality,
  c("Local Authority", "Lung Cancer Registrations per 100,000 (2017-2019)", "Lung Cancer Mortality Rate per 100,000 (2021-2023)")
)

smoking_deprivation <- rename_columns(
  smoking_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)", "Smoking Prevalence (%) (2022-2023)")
)

smoking_lung_cancer_registrations <- rename_columns(
  smoking_lung_cancer_registrations,
  c("Local Authority", "Smoking Prevalence (%) (2022-2023)", "Lung Cancer Registrations per 100,000 (2017-2019)")
)

smoking_lung_cancer_mortality <- rename_columns(
  smoking_lung_cancer_mortality,
  c("Local Authority", "Smoking Prevalence (%) (2022-2023)", "Lung Cancer Mortality Rate per 100,000 (2021-2023)")
)


smoking_deprivation_lung_cancer_registrations <- rename_columns(
  smoking_deprivation_lung_cancer_registrations,
   c("Local Authority", "IMD Deprivation Score (2019)", "Lung Cancer Registrations per 100,000 (2017-2019)", "Smoking Prevalence (%) (2022-2023)")
)

smoking_deprivation_lung_cancer_mortality <- rename_columns(
  smoking_deprivation_lung_cancer_mortality,
   c("Local Authority", "IMD Deprivation Score (2019)", "Smoking Prevalence (%) (2022-2023)", "Lung Cancer Mortality Rate per 100,000 (2021-2023)")
)
  
# Display the Processed Datasets 
lungcancer_registrations_deprivation
lungcancer_mortality_deprivation
lungcancer_registrations_mortality
smoking_deprivation
smoking_lung_cancer_registrations
smoking_deprivation_lung_cancer_mortality
smoking_deprivation_lung_cancer_registrations
smoking_deprivation_lung_cancer_mortality


```

# 2. Lung Cancer Registrations and Deprivation Analysis

## Fit and Plot Linear Regression Model for Lung Cancer Registrations and Deprivation 
```{r}

# Fit the linear regression model to predict Lung Cancer Registrations based on IMD Deprivation Score

lm_lung_cancer_registration_vs_imd <- lm(
  `Lung Cancer Registrations per 100,000 (2017-2019)` ~ `IMD Deprivation Score (2019)`, 
  data = lungcancer_registrations_deprivation
)

# Summarise the linear regression model 
summary(lm_lung_cancer_registration_vs_imd)

# Plot the linear regression model
plot(
  lungcancer_registrations_deprivation$`IMD Deprivation Score (2019)`,
  lungcancer_registrations_deprivation$`Lung Cancer Registrations per 100,000 (2017-2019)`,
  xlab = "IMD Deprivation Score (2019)",
  ylab = "Lung Cancer Registrations per 100,000 (2017–2019)",
  main = "Linear Regression of Lung Cancer Registrations and Deprivation across Lancashire-12",
  pch = 19,
  col = rgb(0, 0, 1, 0.6),  # Blue with 60% opacity
  xlim = c(min(lungcancer_registrations_deprivation$`IMD Deprivation Score (2019)`) * 0.95,
           max(lungcancer_registrations_deprivation$`IMD Deprivation Score (2019)`) * 1.05),
  cex.lab = 1.2,
  cex.main = 1.3
)

# Add red regression line
abline(lm_lung_cancer_registration_vs_imd, col = "red", lwd = 2)



# Add labels for each Local Authority to identify the data points
text(
  lungcancer_registrations_deprivation$`IMD Deprivation Score (2019)`,  
  lungcancer_registrations_deprivation$`Lung Cancer Registrations per 100,000 (2017-2019)`,  
  labels = lungcancer_registrations_deprivation$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)



```

## Lung Cancer Registrations vs Deprivation Linear Regression Diagnostics
```{r}

# Set layout for 2x2 grid of the 4 diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Registrations vs Deprivation model
plot(lm_lung_cancer_registration_vs_imd)


# Calculate Cook's Distance for the Lung Cancer Registrations vs Deprivation model
cooksd <- cooks.distance(lm_lung_cancer_registration_vs_imd)

# Plot Cook's Distance 
plot(
  cooksd,
  main = "Cook's Distance for Lung Cancer Registrations vs Deprivation",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2                   # Line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd),
  y = cooksd,
  labels = lungcancer_registrations_deprivation$`Local Authority`,  # Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)

# Add a threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold


```

# 3. Lung Cancer Mortality Rate and Deprivation Analysis

## Fit and Plot Linear Regression Model for Lung Cancer Mortality Rate and Deprivation 
```{r}

# Fit the linear regression model to predict Lung Cancer Mortality Rate based on IMD Deprivation Score 
lm_lung_cancer_mortality_vs_imd <- lm(
  `Lung Cancer Mortality Rate per 100,000 (2021-2023)` ~ `IMD Deprivation Score (2019)`, 
  data = lungcancer_mortality_deprivation
)

# Summarise the linear regression model 
summary(lm_lung_cancer_mortality_vs_imd)

# Plot the data points and the regression line
plot(
  lungcancer_mortality_deprivation$`IMD Deprivation Score (2019)`,   
  lungcancer_mortality_deprivation$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`,   
  xlab = "IMD Deprivation Score (2019)",   # X-axis label
  ylab = "Lung Cancer Mortality Rate per 100,000 (2021-2023)",   # Y-axis label
  main = "Linear Regression of Lung Cancer Mortality Rate and IMD Deprivation across Lancashire-12",   # Plot title
  pch = 19,   # Solid dots
  col = rgb(0, 0, 1, 0.6)  # Blue with 60% opacity
)

# Add a red regression line to the model
abline(lm_lung_cancer_mortality_vs_imd, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  lungcancer_mortality_deprivation$`IMD Deprivation Score (2019)`,  
  lungcancer_mortality_deprivation$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`,  
  labels = lungcancer_mortality_deprivation$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)


```

## Lung Cancer Mortality Rate vs Deprivation Linear Regression Diagnostics 

```{r}

# Set layout for 2x2 grid of the 4 diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Mortality vs Deprivation model
plot(lm_lung_cancer_mortality_vs_imd)


# Calculate Cook's Distance for the Lung Cancer Mortality vs Deprivation model
cooksd <- cooks.distance(lm_lung_cancer_mortality_vs_imd)

# Plot Cook's Distance 
plot(
  cooksd,
  main = "Cook's Distance for Lung Cancer Mortality vs Deprivation",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2                   # Line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd),
  y = cooksd,
  labels = lungcancer_mortality_deprivation$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)


# Add the threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```
## Quadratic Transformation and Comparison of Linear and Quadratic Regression Models
```{r}
# Fit the linear regression model with a quadratic term
lm_lung_cancer_mortality_vs_imd_quad <- lm(
  `Lung Cancer Mortality Rate per 100,000 (2021-2023)` ~ 
  `IMD Deprivation Score (2019)` + I(`IMD Deprivation Score (2019)`^2),
  data = lungcancer_mortality_deprivation
)

# Summarise the model
summary(lm_lung_cancer_mortality_vs_imd_quad)


# Compare models with ANOVA to assess whether the quadratic term improves the model
anova(lm_lung_cancer_mortality_vs_imd, lm_lung_cancer_mortality_vs_imd_quad)


```

# 4. Lung Cancer Registrations and Mortality Rate Analysis

## Fit and Plot Linear Regression Model for Lung Cancer Registrations and Mortality Rate 
```{r}

# Fit the linear regression model to predict Lung Cancer Mortality Rate based on Cancer Registrations
lm_lung_cancer_mortality_vs_registrations <- lm(
  `Lung Cancer Mortality Rate per 100,000 (2021-2023)` ~ `Lung Cancer Registrations per 100,000 (2017-2019)`,  
  data = lungcancer_registrations_mortality
)

# Summarise the linear regression model
summary(lm_lung_cancer_mortality_vs_registrations)

# Plot the data points and the regression line
plot(
 lungcancer_registrations_mortality$`Lung Cancer Registrations per 100,000 (2017-2019)`,  # X-axis
  lungcancer_registrations_mortality$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`,  # Y-axis
  xlab = "Lung Cancer Registrations per 100,000 (2017-2019)",  # X-axis label
  ylab = "Lung Cancer Mortality Rate per 100,000 (2021-2023)",  # Y-axis label
  main = "Linear Regression of Lung Cancer Mortality Rate and Registrations across Lancashire-12",  # Plot title
  pch = 19,  # Solid dots for data points
  col = rgb(0, 0, 1, 0.6)  # Blue dots with 60% opacity
)

# Add a red regression line to the plot
abline(lm_lung_cancer_mortality_vs_registrations, col = "red")


# Add labels for each Local Authority to identify the data points
text(
  lungcancer_registrations_mortality$`Lung Cancer Registrations per 100,000 (2017-2019)`,  
  lungcancer_registrations_mortality$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`,  
  labels = lungcancer_registrations_mortality$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)



```

## Lung Cancer Registrations vs Mortality Rate Linear Regression Diagnostic Plots
```{r}

# Set layout for 2x2 grid of the 4 diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Mortality vs Registrations model
plot(lm_lung_cancer_mortality_vs_registrations)


# Calculate Cook's Distance for the Lung Cancer Mortality vs Registrations model
cooksd <- cooks.distance(lm_lung_cancer_mortality_vs_registrations)

# Plot Cook's Distance 
plot(
  cooksd,
  main = "Cook's Distance for Lung Cancer Mortality vs Registrations",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2                   # Line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd),
  y = cooksd,
  labels = lungcancer_registrations_mortality$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)


# Add the threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

# 5. Smoking and Deprivation Analysis

## Fit and Plot Linear Regression Model for Smoking Prevalence and Deprivation
```{r}

# Fit the linear regression model to predict Smoking Prevalence based on IMD Deprivation Score
lm_smoking_prevalence_vs_deprivation <- lm(
  `Smoking Prevalence (%) (2022-2023)` ~ `IMD Deprivation Score (2019)`, 
  data = smoking_deprivation
)

# Summarise the linear regression model
summary(lm_smoking_prevalence_vs_deprivation)

# Plot the data points and the regression line
plot(
  smoking_deprivation$`IMD Deprivation Score (2019)`, 
  smoking_deprivation$`Smoking Prevalence (%) (2022-2023)`, 
  xlab = "IMD Deprivation Score (2019)",  # X-axis label
  ylab = "Smoking Prevalence (%) (2022-2023)",  # Y-axis label
  main = "Linear Regression of Smoking Prevalence and IMD Deprivation Score across Lancashire-12",  # Plot title
  pch = 19,  # Solid dots for data points
  col = rgb(0, 0, 1, 0.6)  # Blue dots with 60% opacity
)

# Add a red regression line to the plot
abline(lm_smoking_prevalence_vs_deprivation, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  smoking_deprivation$`IMD Deprivation Score (2019)`,  
  smoking_deprivation$`Smoking Prevalence (%) (2022-2023)`,  
  labels = smoking_deprivation$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)



```

## Smoking Prevalence vs Deprivation Linear Regression Diagnostics

```{r}

# Set 2x2 grid for diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Mortality vs Deprivation model
plot(lm_smoking_prevalence_vs_deprivation)


# Calculate Cook's Distance for the Lung Cancer Mortality vs Deprivation model
cooksd <- cooks.distance(lm_smoking_prevalence_vs_deprivation)

# Plot Cook's Distance
plot(
  cooksd,
  main = "Cook's Distance for Smoking Prevalence vs Deprivation",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2,                  # Line width
  ylim = c(0, 0.35)         # Extend the upper limit of the y-axis to 0.35 so threshold line is visible
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd),
  y = cooksd,
  labels = smoking_deprivation$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)


# Add the threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

# 6. Smoking and Lung Cancer Registrations Analysis

## Fit and Plot Linear Regression Model for Smoking Prevalence and Lung Cancer Registrations
```{r}
# Fit the linear regression model to predict Lung Cancer Registrations based on Smoking Prevalence
lm_lung_cancer_registrations_vs_smoking <- lm(
  `Lung Cancer Registrations per 100,000 (2017-2019)` ~ `Smoking Prevalence (%) (2022-2023)`, 
  data = smoking_lung_cancer_registrations
)

# Summarise the linear regression model
summary(lm_lung_cancer_registrations_vs_smoking)

# Plot the data points and the regression line
plot(
  smoking_lung_cancer_registrations$`Smoking Prevalence (%) (2022-2023)`, 
  smoking_lung_cancer_registrations$`Lung Cancer Registrations per 100,000 (2017-2019)`, 
  xlab = "Smoking Prevalence (%) (2022-2023)",  # X-axis label
  ylab = "Lung Cancer Registrations per 100,000 (2017-2019)",  # Y-axis label
  main = "Linear Regression of Lung Cancer Registrations and Smoking across Lancashire-12",  # Plot title
  pch = 19,  # Solid dots for data points
  col = rgb(0, 0, 1, 0.6)  # Blue dots with 60% opacity
)

# Add a red regression line to the plot
abline(lm_lung_cancer_registrations_vs_smoking, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  smoking_lung_cancer_registrations$`Smoking Prevalence (%) (2022-2023)`,  
  smoking_lung_cancer_registrations$`Lung Cancer Registrations per 100,000 (2017-2019)`,  
  labels = smoking_lung_cancer_registrations$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)




```

## Smoking Prevalence vs Lung Cancer Registrations Linear Regression Diagnostic Plots
```{r}

# Set 2x2 grid for diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Incidence vs Smoking model
plot(lm_lung_cancer_registrations_vs_smoking)


# Calculate Cook's Distance for the Lung Cancer Incidence vs Smoking model
cooksd <- cooks.distance(lm_lung_cancer_registrations_vs_smoking)

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  main = "Cook's Distance for Lung Cancer Incidence vs Smoking",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2,                  # Line width
  ylim = c(0, 0.5)          # Extend the upper limit of the y-axis to 0.5 so threshold line is visible
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd),
  y = cooksd,
  labels = smoking_lung_cancer_registrations$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)


# Add the threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold
  


```


# 7. Smoking and Lung Cancer Mortality Analysis

## Fit and Plot Linear Regression Model for Smoking Prevalence and Lung Cancer Mortality
```{r}

# Fit the linear regression model to predict Lung Cancer Mortality based on Smoking Prevalence
lm_lung_cancer_mortality_vs_smoking <- lm(
  `Lung Cancer Mortality Rate per 100,000 (2021-2023)` ~ `Smoking Prevalence (%) (2022-2023)`, 
  data = smoking_lung_cancer_mortality
)

# Summarise the linear regression model
summary(lm_lung_cancer_mortality_vs_smoking)

# Plot the data points and the regression line
plot(
  smoking_lung_cancer_mortality$`Smoking Prevalence (%) (2022-2023)`, 
  smoking_lung_cancer_mortality$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`, 
  xlab = "Smoking Prevalence (%) (2022-2023)",  # X-axis label
  ylab = "Lung Cancer Mortality Rate per 100,000 (2021-2023)",  # Y-axis label
  main = "Linear Regression of Lung Cancer Mortality vs Smoking Prevalence",  # Plot title
  pch = 19,  # Solid dots for data points
  col = rgb(0, 0, 1, 0.6)  # Blue dots with 60% opacity
)

# Add a red regression line to the plot
abline(lm_lung_cancer_mortality_vs_smoking, col = "red")


# Add labels for each Local Authority to identify the data points
text(
  smoking_lung_cancer_mortality$`Smoking Prevalence (%) (2022-2023)`,  
  smoking_lung_cancer_mortality$`Lung Cancer Mortality Rate per 100,000 (2021-2023)`,  
  labels = smoking_lung_cancer_mortality$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)



```
## Smoking Prevalence vs Lung Cancer Mortality Linear Regression Diagnostic Plots 
```{r}

# Set 2x2 grid for diagnostic plots
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Lung Cancer Mortality and Smoking model
plot(lm_lung_cancer_mortality_vs_smoking)

# Calculate Cook's Distance for the Lung Cancer Mortality and Smoking model
cooksd_mortality <- cooks.distance(lm_lung_cancer_mortality_vs_smoking)

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd_mortality,
  main = "Cook's Distance - Lung Cancer Mortality Model",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2,                  # Line width
  ylim = c(0, 0.5)          # Extend the upper limit of the y-axis
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd_mortality),
  y = cooksd_mortality,
  labels = smoking_lung_cancer_mortality$`Local Authority`,  # Local Authority names
  cex = 0.7,                # Adjust text size for readability
  pos = 4,                  # Position labels to the right of each point
  col = "blue"              # Set label colour to blue
)

# Add the threshold line for Cook's Distance (4/n rule)
threshold_mortality <- 4 / length(cooksd_mortality)  # Common threshold is 4/n
abline(h = threshold_mortality, col = "purple", lty = 2)  # Dashed purple line for threshold


```

# 8. Smoking, Deprivation and Lung Cancer Registrations  

## Fit Multiple Linear Regression Model for Smoking, Deprivation and Lung Cancer Registrations with interaction
```{r}
# Fit Multiple Linear Regression Model for Smoking and Deprivation Effect on Lung Cancer Registrations
lm_lungcancer_deprivation_smoking_interaction <- lm(
  `Lung Cancer Registrations per 100,000 (2017-2019)` ~ 
    `IMD Deprivation Score (2019)` * `Smoking Prevalence (%) (2022-2023)`,
  data = smoking_deprivation_lung_cancer_registrations
)

# Summarise the model
summary(lm_lungcancer_deprivation_smoking_interaction)

# Check for multicollinearity using Variance Inflation Factors (VIFs)
vif(lm_lungcancer_deprivation_smoking_interaction)

```
```{r}
# Centre the predictors for lung cancer model as high multicollinearity
smoking_deprivation_lung_cancer_registrations$IMD_centered <- scale(
  smoking_deprivation_lung_cancer_registrations$`IMD Deprivation Score (2019)`,
  center = TRUE, scale = FALSE
)

smoking_deprivation_lung_cancer_registrations$Smoking_centered <- scale(
  smoking_deprivation_lung_cancer_registrations$`Smoking Prevalence (%) (2022-2023)`,
  center = TRUE, scale = FALSE
)

# Fit the multiple regression model with centered predictors and interaction
lm_lungcancer_deprivation_smoking_interaction_centered <- lm(
  `Lung Cancer Registrations per 100,000 (2017-2019)` ~ 
    IMD_centered * Smoking_centered,
  data = smoking_deprivation_lung_cancer_registrations
)

# Summarise the centred model
summary(lm_lungcancer_deprivation_smoking_interaction_centered)

# Check VIFs after centring
vif(lm_lungcancer_deprivation_smoking_interaction_centered)

```

```

