---
title: "Bowel_Cancer_Code"
output: html_document
date: "2025-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Set Up

## Install and Load Packages
```{r}

library(tidyverse) # load tidyverse package to make graphs and format the data in tables
library(car) # load car package for vifs to test for multicollinearity in multiple linear regressions
library(effects) # load effects package to to calculate the marginal effects of a variable in multiple regression models to plot relationships 

```

## Load Data for the Bowel Cancer Analysis from the Main Project Folder
```{r}

# Load the CSV files for Bowel Cancer Incidence vs Deprivation
bowelcancer_incidence_deprivation <- read.csv("bowel_cancer_incidence_deprivation.csv")
bowelcancer_incidence_deprivation

# Load the CSV files for Bowel Cancer Mortality vs Deprivation
bowelcancer_mortality_deprivation <- read.csv("bowel_cancer_mortality_deprivation.csv")
bowelcancer_mortality_deprivation

# Load the CSV files for Bowel Cancer Mortality vs Incidence
bowelcancer_incidence_mortality <- read.csv("bowel_cancer_incidence_mortality.csv")
bowelcancer_incidence_mortality

# Load the CSV files for Bowel Cancer Deprivation vs Screening 
bowelcancer_screening_deprivation <- read.csv("bowel_cancer_screening_deprivation.csv")
bowelcancer_screening_deprivation

# Load the CSV files for Bowel Cancer Screening vs Incidence

bowelcancer_screening_incidence <- read.csv("bowel_cancer_screening_incidence.csv")
bowelcancer_screening_incidence


# Load the CSV files for Bowel Cancer Screening vs Mortality

bowelcancer_screening_mortality <- read.csv("bowel_cancer_screening_mortality .csv")
bowelcancer_screening_mortality

# Load the CSV files for Bowel Cancer Screening + Deprivation on Incidence

bowelcancer_screening_deprivation_incidence <- read.csv("bowelcancer_screening_deprivation_incidence.csv")
bowelcancer_screening_deprivation_incidence

# Load the CSV files for Bowel Cancer Screening + Deprivation on Mortality

bowelcancer_screening_deprivation_mortality <- read.csv("bowel_cancer_screening_deprivation_mortality.csv")
bowelcancer_screening_deprivation_mortality




```


## Data Processing to ensure uniformity across column names in datasets
```{r}


# Function to rename the columns in a dataset  
rename_columns <- function(dataset, new_colnames) {  
   
   # Check if the number of new column names matches the number of columns in the dataset  
   if (ncol(dataset) != length(new_colnames)) {  
     stop("Error: The number of columns in the dataset does not match the provided column names.")  
     # stop() halts execution and returns an error message if the column counts do not match  
   }  
   
   # Assign the new column names to the dataset  
   colnames(dataset) <- new_colnames  
   
   # Return the modified dataset with the updated column names  
   return(dataset)  
}  

# Rename Columns for Each Bowel Cancer Dataset
bowelcancer_incidence_deprivation <- rename_columns(
  bowelcancer_incidence_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)", "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)")
)

bowelcancer_mortality_deprivation <- rename_columns(
  bowelcancer_mortality_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)", "Bowel Cancer Mortality Rate per 100,000 (2021-2023)")
)

bowelcancer_incidence_mortality <- rename_columns(
  bowelcancer_incidence_mortality,
  c("Local Authority", "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)", "Bowel Cancer Mortality Rate per 100,000 (2021-2023)")
)


bowelcancer_screening_deprivation <- rename_columns(
  bowelcancer_screening_deprivation,
  c("Local Authority", "IMD Deprivation Score (2019)","Bowel Cancer Screening Coverage (2023) (%)")
)


bowelcancer_screening_incidence <- rename_columns(
  bowelcancer_screening_incidence,
  c("Local Authority", "Bowel Cancer Screening Coverage (2023) (%)", "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)")
)

bowelcancer_screening_deprivation_incidence <- rename_columns(
  bowelcancer_screening_deprivation_incidence,
  c("Local Authority", "Bowel Cancer Screening Coverage (2023) (%)", "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)","IMD Deprivation Score (2019)" )
)

bowelcancer_screening_mortality <- rename_columns(
  bowelcancer_screening_mortality,
  c("Local Authority", "Bowel Cancer Screening Coverage (2023) (%)", "Bowel Cancer Mortality Rate per 100,000 (2021-2023)")
)

bowelcancer_screening_deprivation_mortality <- rename_columns(
  bowelcancer_screening_deprivation_mortality,
  c("Local Authority", "Bowel Cancer Screening Coverage (2023) (%)", "Bowel Cancer Mortality Rate per 100,000 (2021-2023)","IMD Deprivation Score (2019)" )
)


# Display the Processed Datasets
bowelcancer_incidence_deprivation
bowelcancer_mortality_deprivation
bowelcancer_incidence_mortality
bowelcancer_screening_deprivation
bowelcancer_screening_incidence
bowelcancer_screening_deprivation_incidence
bowelcancer_screening_mortality
bowelcancer_screening_deprivation_mortality 


```

# 2. Bowel Cancer Incidence and Deprivation Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Incidence and Deprivation 
```{r}
# Fit the linear regression model to predict Bowel Cancer Incidence based on IMD Deprivation Score
lm_bowel_cancer_incidence_vs_imd <- lm(
  `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)` ~ `IMD Deprivation Score (2019)`,
  data = bowelcancer_incidence_deprivation
)

# Summarise the linear regression model
summary(lm_bowel_cancer_incidence_vs_imd)

# Plot the data points and the regression line
plot(
  bowelcancer_incidence_deprivation$`IMD Deprivation Score (2019)`,
  bowelcancer_incidence_deprivation$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,
  xlab = "IMD Deprivation Score (2019)",  # Add the X-axis label
  ylab = "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)",  # Add the Y-axis label
  main = "Linear Regression of Bowel Cancer Incidence vs IMD Deprivation",  # Add plot title
  pch = 19,  # Add solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Make the dots blue with 60% opacity
  xlim = c(
    min(bowelcancer_incidence_deprivation$`IMD Deprivation Score (2019)`),
    max(bowelcancer_incidence_deprivation$`IMD Deprivation Score (2019)`) * 1.1
  )  # Extend the x-axis to fit all the points
)

# Add a red regression line to the model
abline(lm_bowel_cancer_incidence_vs_imd, col = "red")


# Add labels for each Local Authority to identify the data points
text(
  bowelcancer_incidence_deprivation$`IMD Deprivation Score (2019)`,  
   bowelcancer_incidence_deprivation$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,  
  labels =  bowelcancer_incidence_deprivation$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)




```
## Bowel Cancer Incidence vs Deprivation Linear Regression Diagnosis
```{r}
# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for the Bowel Cancer Incidence vs Deprivation model
plot(lm_bowel_cancer_incidence_vs_imd)

# Calculate Cook's Distance for the Bowel Cancer Incidence model
cooksd <- cooks.distance(lm_bowel_cancer_incidence_vs_imd)

# Plot Cook's Distance
plot(
  cooksd,
  main = "Cook's Distance for Bowel Cancer Incidence vs Deprivation",
  ylab = "Cook's Distance",
  type = "h",          # Use vertical lines for Cook's Distance
  col = "red",         # Make the lines red
  lwd = 2              # Adjust the line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_incidence_deprivation$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add a threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold


```
# 3. Bowel Cancer Mortality Rate and Deprivation Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Mortality Rate and Deprivation 
```{r}

# Fit the linear regression model to predict Bowel Cancer Mortality based on IMD Deprivation Score
lm_bowel_cancer_mortality_vs_imd <- lm(
  `Bowel Cancer Mortality Rate per 100,000 (2021-2023)` ~ `IMD Deprivation Score (2019)`,
  data = bowelcancer_mortality_deprivation
)

# Summarise the linear regression model
summary(lm_bowel_cancer_mortality_vs_imd)

# Plot the data points and the regression line
plot(
  bowelcancer_mortality_deprivation$`IMD Deprivation Score (2019)`,
  bowelcancer_mortality_deprivation$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,
  xlab = "IMD Deprivation Score (2019)",  # Add the X-axis label
  ylab = "Bowel Cancer Mortality Rate per 100,000 (2021-2023)",  # Add the Y-axis label
  main = "Linear Regression of Bowel Cancer Mortality vs IMD Deprivation",  # Add plot title
  pch = 19,  # Solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Blue dots with 60% opacity
  xlim = c(
    min(bowelcancer_mortality_deprivation$`IMD Deprivation Score (2019)`),
    max(bowelcancer_mortality_deprivation$`IMD Deprivation Score (2019)`) * 1.1
  )  # Extend the x-axis to fit all the points
)

# Add a red regression line to the model
abline(lm_bowel_cancer_mortality_vs_imd, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  bowelcancer_mortality_deprivation$`IMD Deprivation Score (2019)`,  # Extract the X-axis coordinates
  bowelcancer_mortality_deprivation$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,  # Extract the Y-axis coordinates
  labels = bowelcancer_mortality_deprivation$`Local Authority`,  # Add the Local Authority names as labels
  pos = 4,  # Position the labels to the right of each point
  cex = 0.6,  # Adjust text size for readability
  col = "black"  # Set the text color to black for the labels
)


```

## Bowel Cancer Mortality Rate vs Deprivation Linear Regression Diagnostics

```{r}
# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for the Bowel Cancer Mortality vs Deprivation model
plot(lm_bowel_cancer_mortality_vs_imd)

# Calculate Cook's Distance for the Bowel Cancer Mortality model
cooksd <- cooks.distance(lm_bowel_cancer_mortality_vs_imd)

# Plot Cook's Distance
plot(
  cooksd,
  main = "Cook's Distance for Bowel Cancer Mortality vs Deprivation",
  ylab = "Cook's Distance",
  type = "h",          # Use vertical lines for Cook's Distance
  col = "red",         # Make the lines red
  lwd = 2              # Adjust the line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_mortality_deprivation$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add a threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold


```


# 4. Bowel Cancer (SIR) and Mortality Rate Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Incidence and Mortality Rate
```{r}
## Fit the linear regression model to predict Bowel Cancer Mortality based on Incidence
lm_bowel_cancer_incidence_vs_mortality <- lm(
  `Bowel Cancer Mortality Rate per 100,000 (2021-2023)` ~ `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,
  data = bowelcancer_incidence_mortality
)

# Summarise the linear regression model
summary(lm_bowel_cancer_incidence_vs_mortality)

# Plot the data points and the regression line
plot(
  bowelcancer_incidence_mortality$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,
  bowelcancer_incidence_mortality$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,
  xlab = "Incidence of Bowel Cancer (SIR) per 100 (2015-2019)",  # X-axis label
  ylab = "Bowel Cancer Mortality Rate per 100,000 (2021-2023)",  # Y-axis label
  main = "Linear Regression of Bowel Cancer Mortality vs Incidence",  # Plot title
  pch = 19,  # Solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Blue dots with 60% opacity
  xlim = c(
    min(bowelcancer_incidence_mortality$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`),
    max(bowelcancer_incidence_mortality$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`) * 1.1
  )  # Extend the x-axis to fit all the points
)

# Add a red regression line to the model
abline(lm_bowel_cancer_incidence_vs_mortality, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  bowelcancer_incidence_mortality$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,  
  bowelcancer_incidence_mortality$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,  
  labels = bowelcancer_incidence_mortality$`Local Authority`,  
  pos = 4,  
  cex = 0.6,  
  col = "black"  
)

```

## Bowel Cancer Mortality Rate vs Incidence Linear Regression Diagnostics
```{r}
# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for the Bowel Cancer Incidence and Mortality model
plot(lm_bowel_cancer_incidence_vs_mortality)

# Calculate Cook's Distance for the Bowel Cancer Incidence and Mortality model
cooksd <- cooks.distance(lm_bowel_cancer_incidence_vs_mortality)

# Compute the threshold for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  main = "Cook's Distance for Bowel Cancer Incidence vs Mortality",
  ylab = "Cook's Distance",
  type = "h",          # Use vertical lines for Cook's Distance
  col = "red",         # Make the lines red
  lwd = 2,             # Adjust the line width
  ylim = c(0, max(cooksd, threshold) * 1.2)  # Extend y-axis to fit the threshold line
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_incidence_mortality$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add the threshold line for Cook's Distance
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

# 5. Bowel Cancer Screening and Deprivation Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Screening and Deprivation

```{r}
## Fit the linear regression model to predict Bowel Cancer Mortality based on Incidence
lm_bowel_cancer_screening_vs_imd <- lm(
  `Bowel Cancer Screening Coverage (2023) (%)` ~ `IMD Deprivation Score (2019)`,
  data = bowelcancer_screening_deprivation
)

# Summarise the linear regression model
summary(lm_bowel_cancer_screening_vs_imd)

# Plot the data points and the regression line
plot(
 bowelcancer_screening_deprivation$`IMD Deprivation Score (2019)`,
 bowelcancer_screening_deprivation$`Bowel Cancer Screening Coverage (2023) (%)`,
  xlab = "IMD Deprivation Score (2019)",  # X-axis label
  ylab = "Bowel Cancer Screening Coverage (2023) (%)",  # Y-axis label
  main = "Linear Regression of Bowel Cancer Screening Coverage vs IMD Deprivation",  # Plot title
  pch = 19,  # Solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Blue dots with 60% opacity
  xlim = c(
    min( bowelcancer_screening_deprivation$`IMD Deprivation Score (2019)`),
    max( bowelcancer_screening_deprivation$`IMD Deprivation Score (2019)`) * 1.1
  )  # Extend the x-axis to fit all the points
)

# Add a red regression line to the model
abline(lm_bowel_cancer_screening_vs_imd, col = "red")

# Add labels for each Local Authority to identify the data points
text(
  bowelcancer_screening_deprivation$`IMD Deprivation Score (2019)`,  # X-axis coordinates
  bowelcancer_screening_deprivation$`Bowel Cancer Screening Coverage (2023) (%)`,  # Y-axis coordinates
  labels = bowelcancer_screening_deprivation$`Local Authority`,  # Local Authority names
  pos = 4,  # Position labels to the right of each point
  cex = 0.6,  # Adjust text size for readability
  col = "black"  # Set the text color to black for the labels
)

```

## Bowel Cancer Screening Coverage vs Deprivation Linear Regression Diagnostics

```{r}
# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2))          # Arrange plots in a 2x2 grid

# Create the 4 diagnostic plots for the Bowel Cancer Screening vs Deprivation model
plot(lm_bowel_cancer_screening_vs_imd)

# Calculate Cook's Distance for the Bowel Cancer Screening Coverage and Deprivation model
cooksd <- cooks.distance(lm_bowel_cancer_screening_vs_imd)

# Compute the threshold for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  main = "Cook's Distance for Bowel Cancer Screening Coverage vs Deprivation",
  ylab = "Cook's Distance",
  type = "h",          # Use vertical lines for Cook's Distance
  col = "red",         # Make the lines red
  lwd = 2,             # Adjust the line width
  ylim = c(0, max(cooksd, threshold) * 1.2)  # Extend y-axis to fit the threshold line
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_screening_deprivation$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add the threshold line for Cook's Distance
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

## Sensitivity Analysis without Burnley
```{r}
# Filter out Burnley from the dataset to create a filtered dataset
filtered_bowel_screening_deprivation <- subset(
  bowelcancer_screening_deprivation,
  `Local Authority` != "Burnley"
)

# Fit the linear regression model on the filtered dataset
lm_bowel_screening_vs_imd_filtered <- lm(
  `Bowel Cancer Screening Coverage (2023) (%)` ~ `IMD Deprivation Score (2019)`,
  data = filtered_bowel_screening_deprivation
)

# Summarise the linear regression model
summary(lm_bowel_screening_vs_imd_filtered)

# Plot the data points and the regression line
plot(
  filtered_bowel_screening_deprivation$`IMD Deprivation Score (2019)`,
  filtered_bowel_screening_deprivation$`Bowel Cancer Screening Coverage (2023) (%)`,
  xlab = "IMD Deprivation Score (2019)",  # X-axis label
  ylab = "Bowel Cancer Screening Coverage (2023) (%)",  # Y-axis label
  main = "Bowel Cancer Screening vs IMD Deprivation (Excluding Burnley)",  # Plot title
  pch = 19,  # Solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Blue dots with 60% opacity
  xlim = c(
    min(filtered_bowel_screening_deprivation$`IMD Deprivation Score (2019)`),
    max(filtered_bowel_screening_deprivation$`IMD Deprivation Score (2019)`) * 1.1
  )  # Extend the x-axis to fit all the points
)

# Add the regression line
abline(lm_bowel_screening_vs_imd_filtered, col = "red")


# Add labels for each Local Authority to the plot
text(
  filtered_bowel_screening_deprivation$`IMD Deprivation Score (2019)`,
  filtered_bowel_screening_deprivation$`Bowel Cancer Screening Coverage (2023) (%)`,
  labels = filtered_bowel_screening_deprivation$`Local Authority`,
  pos = 4,  # Position the text to the right of the data points
  cex = 0.6,  # Adjust the text size
  col = "black"  # Colour of the labels
)


```




# 6. Bowel Cancer Screening and Incidence Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Screening and Incidence
```{r}
# Fit linear regression model
lm_screening_vs_incidence <- lm(
  `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)` ~ `Bowel Cancer Screening Coverage (2023) (%)`,
  data = bowelcancer_screening_incidence
)

# Summary of model
summary(lm_screening_vs_incidence)

# Scatterplot of the linear regression model 
plot(
  bowelcancer_screening_incidence$`Bowel Cancer Screening Coverage (2023) (%)`,
  bowelcancer_screening_incidence$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,
  xlab = "Bowel Cancer Screening Coverage (2023) (%)",
  ylab = "Bowel Cancer Incidence (SIR) per 100 (2015–2019)",
  main = "Regression: Incidence vs Screening Coverage",
  pch = 19,
  col = rgb(0, 0, 1, 0.6)
)

# Add regression line
abline(lm_screening_vs_incidence, col = "red")


# Add local authority labels
text(
  bowelcancer_screening_incidence$`Bowel Cancer Screening Coverage (2023) (%)`,
  bowelcancer_screening_incidence$`Incidence of Bowel Cancer (SIR) per 100 (2015-2019)`,
  labels = bowelcancer_screening_incidence$`Local Authority`,
  pos = 4,
  cex = 0.6,
  col = "black"
)
```

## Bowel Cancer Screening Coverage vs Incidence Linear Regression Diagnostics
```{r}

# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for the Screening vs Incidence model
plot(lm_screening_vs_incidence)


# Calculate Cook's Distance for the model
cooksd <- cooks.distance(lm_screening_vs_incidence)

# Compute the threshold for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  type = "h",           # Use vertical lines for Cook's Distance
  col = "red",          # Make the lines red
  lwd = 2,              # Line width
  ylim = c(0, max(cooksd, threshold) * 1.2),  # Extend y-axis to fit the threshold line
  main = "Cook's Distance: Screening vs Incidence",  # Plot title
  ylab = "Cook's Distance"  # Y-axis label
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_screening_incidence$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add the threshold line for Cook's Distance
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold


```


# 7. Bowel Cancer Screening, Deprivation and Incidence Analysis

## Fit Multiple Linear Regression Regression Model with interaction term of deprivation and bowel cancer screening coverage on  Incidence

```{r}
# Fit the multiple regression model with interaction between deprivation and screening
lm_incidence_deprivation_screening_interaction <- lm(
  `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)` ~ 
    `IMD Deprivation Score (2019)` * `Bowel Cancer Screening Coverage (2023) (%)`,
  data = bowelcancer_screening_deprivation_incidence
)

# Summarise the model
summary(lm_incidence_deprivation_screening_interaction)
```

## Check for Multicollinarity with VIF values for this model 

```{r}

# Calculate VIF
vif(lm_incidence_deprivation_screening_interaction)  # Replace lm_bowel with your model name if different
```

## As the VIF values are above 10, centre the variables and refit the multiple linear regression model 
```{r}
# Centre the IMD Deprivation Score and Screening Coverage variables
bowelcancer_screening_deprivation_incidence$IMD_centered <- scale(
  bowelcancer_screening_deprivation_incidence$`IMD Deprivation Score (2019)`, 
  center = TRUE, scale = FALSE
)

bowelcancer_screening_deprivation_incidence$Screening_centered <- scale(
  bowelcancer_screening_deprivation_incidence$`Bowel Cancer Screening Coverage (2023) (%)`, 
  center = TRUE, scale = FALSE
)

# Refit the model using centred variables
lm_incidence_deprivation_screening_centered <- lm(
  `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)` ~ 
    IMD_centered * Screening_centered,
  data = bowelcancer_screening_deprivation_incidence
)

# Summarise the centred model
summary(lm_incidence_deprivation_screening_centered)
```

## Check VIFs of centred model
```{r}

# Check VIFs
vif(lm_incidence_deprivation_screening_centered)

```



## Bowel Cancer Screening Coverage, Deprivation Bowel Cancer Incidence  Regression Diagnostics

```{r}
# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for the Bowel Cancer Screening, Deprivation, and Mortality model
plot(lm_incidence_deprivation_screening_centered)

# Calculate Cook's Distance for the Bowel Cancer Screening, Deprivation, and Mortality model
cooksd <- cooks.distance(lm_incidence_deprivation_screening_centered)

# Compute the threshold for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  main = "Cook's Distance for Bowel Cancer Screening Coverage, Deprivation and Mortality",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2,                  # Line width
  ylim = c(0, max(cooksd, threshold) * 1.2)  # Extend y-axis to fit the threshold line
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_screening_deprivation_incidence$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                 # Adjust text size for readability
  pos = 4,                   # Position labels to the right of each point
  col = "blue"               # Set label colour to blue
)

# Add the threshold line for Cook's Distance
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

## Sensitivity Analysis excluding Burnley
```{r}
# Create a filtered dataset removing Burnley
filtered_bowel_incidence_multiple_regression_model <- subset(
  bowelcancer_screening_deprivation_incidence,
  `Local Authority` != "Burnley"
)

# Re-center the predictors in the filtered dataset
filtered_bowel_incidence_multiple_regression_model$IMD_centered <- scale(
  filtered_bowel_incidence_multiple_regression_model$`IMD Deprivation Score (2019)`,
  center = TRUE,
  scale = FALSE  # Only centre, no scaling
)

filtered_bowel_incidence_multiple_regression_model$Screening_centered <- scale(
  filtered_bowel_incidence_multiple_regression_model$`Bowel Cancer Screening Coverage (2023) (%)`,
  center = TRUE,
  scale = FALSE  # Only centre, no scaling
)

# Refit the multiple linear regression model using the filtered data
lm_bowel_incidence_multiple_regression_model <- lm(
  `Incidence of Bowel Cancer (SIR) per 100 (2015-2019)` ~
    IMD_centered * Screening_centered,
  data = filtered_bowel_incidence_multiple_regression_model
)

# View the model summary
summary(lm_bowel_incidence_multiple_regression_model)

# Check for multicollinearity using VIFs
vif(lm_bowel_incidence_multiple_regression_model)


```

# 8. Bowel Cancer Screening and Mortality Analysis

## Fit and Plot Linear Regression Model for Bowel Cancer Screening and Mortality
```{r}
# Fit the linear regression model to predict Bowel Cancer Mortality from Screening Coverage
lm_bowel_mortality_vs_screening <- lm(
  `Bowel Cancer Mortality Rate per 100,000 (2021-2023)` ~ `Bowel Cancer Screening Coverage (2023) (%)`,
  data = bowelcancer_screening_mortality
)

# Summarise the linear regression model
summary(lm_bowel_mortality_vs_screening)

# Plot the data points and the regression line
plot(
  bowelcancer_screening_mortality$`Bowel Cancer Screening Coverage (2023) (%)`,
  bowelcancer_screening_mortality$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,
  xlab = "Bowel Cancer Screening Coverage (2023) (%)",  # X-axis label
  ylab = "Bowel Cancer Mortality Rate per 100,000 (2021-2023)",  # Y-axis label
  main = "Linear Regression of Bowel Cancer Mortality vs Screening Coverage",  # Plot title
  pch = 19,  # Solid dots for the data points
  col = rgb(0, 0, 1, 0.6),  # Blue dots with 60% opacity
  xlim = c(
    min(bowelcancer_screening_mortality$`Bowel Cancer Screening Coverage (2023) (%)`),
    max(bowelcancer_screening_mortality$`Bowel Cancer Screening Coverage (2023) (%)`) * 1.05
  )  # Extend the x-axis to fit all the points
)

# Add the regression line in red
abline(lm_bowel_mortality_vs_screening, col = "red")

# Add Local Authority labels
text(
  bowelcancer_screening_mortality$`Bowel Cancer Screening Coverage (2023) (%)`,
  bowelcancer_screening_mortality$`Bowel Cancer Mortality Rate per 100,000 (2021-2023)`,
  labels = bowelcancer_screening_mortality$`Local Authority`,
  pos = 4,  # Position labels to the right of each point
  cex = 0.6,  # Adjust text size for readability
  col = "black"  # Set the text colour to black for the labels
)


```

## Diagnostics for Bowel Cancer Screening and Mortality

```{r}

# Set layout for 2x2 grid of diagnostic plots
par(mfrow = c(2, 2))  # Set plotting area to 2x2

# Create the 4 diagnostic plots for the Bowel Cancer Mortality vs Screening model
plot(lm_bowel_mortality_vs_screening)

# Calculate Cook's Distance for the Bowel Cancer Mortality vs Screening model
cooksd <- cooks.distance(lm_bowel_mortality_vs_screening)

# Compute the threshold for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n

# Plot Cook's Distance with an extended y-axis limit
plot(
  cooksd,
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2,                  # Adjust the line width
  ylim = c(0, max(cooksd, threshold) * 1.2),  # Extend y-axis to fit the threshold line
  main = "Cook's Distance: Mortality vs Screening",  # Plot title
  ylab = "Cook's Distance"  # Y-axis label
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_screening_mortality$`Local Authority`,  # Use Local Authority names
  cex = 0.7,             # Adjust text size for readability
  pos = 4,               # Position labels to the right of each point
  col = "blue"           # Set label colour to blue
)

# Add the threshold line for Cook's Distance
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```

# 9. Bowel Cancer Screening, Deprivation and Mortality Analysis

## Fit Multiple Linear Regression Regression Model with interaction term of deprivation and bowel cancer screening coveage on mortality
```{r}
# Fit the multiple regression model with interaction between deprivation and screening
lm_mortality_deprivation_screening_interaction <- lm(
  `Bowel Cancer Mortality Rate per 100,000 (2021-2023)` ~
    `IMD Deprivation Score (2019)` * `Bowel Cancer Screening Coverage (2023) (%)`,
  data = bowelcancer_screening_deprivation_mortality
)

# Summarise the model
summary(lm_mortality_deprivation_screening_interaction)

```

## Check for Multicollinarity with VIF values for this model 
```{r}

# Check for multicollinearity using Variance Inflation Factors (VIFs) 
vif(lm_mortality_deprivation_screening_interaction)

```

## As the VIF values are above 10, centre the variables and refit the multiple linear regression model 
```{r}

# Centre the IMD Deprivation Score and Screening Coverage variables
bowelcancer_screening_deprivation_mortality$IMD_centered <- scale(
  bowelcancer_screening_deprivation_mortality$`IMD Deprivation Score (2019)`,
  center = TRUE, scale = FALSE
)

bowelcancer_screening_deprivation_mortality$Screening_centered <- scale(
  bowelcancer_screening_deprivation_mortality$`Bowel Cancer Screening Coverage (2023) (%)`,
  center = TRUE, scale = FALSE
)

# Refit the model using centred variables
lm_mortality_centered <- lm(
  `Bowel Cancer Mortality Rate per 100,000 (2021-2023)` ~
    IMD_centered * Screening_centered,
  data = bowelcancer_screening_deprivation_mortality
)

# Summarise the centred model
summary(lm_mortality_centered)

# Check VIFs again in centred model
vif(lm_mortality_centered)
```

## Plot the Bowel Cancer Screening Coverage, Deprivation Bowel Cancer Mortality  Regression 
```{r}


# Calculate the marginal effect of the 'Screening' variable (centered) on Bowel Cancer Mortality

effect_screening <- Effect("Screening_centered", lm_mortality_centered)

#  Convert this effect into a dataframe for easier plotting

df_effect <- as.data.frame(effect_screening)

# Plot the predicted relationship between Screening Coverage and Mortality 
ggplot(df_effect, aes(x = Screening_centered, y = fit)) +  #
  geom_line(color = "red", size = 1) +  # Add a red line showing the predicted fit
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey70", alpha = 0.3) +  # Add a shaded confidence interval
  labs(
    title = "Predicted Bowel Cancer Mortality by Screening Coverage (adjusted for deprivation)",  # Title
    x = "Screening Coverage (centered)",  # X-axis label
    y = "Predicted Mortality Rate per 100,000"  # Y-axis label
  ) +
  theme_minimal(base_size = 13) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")  
  )


```



## Bowel Cancer Screening Coverage, Deprivation Bowel Cancer Mortality  Regression Diagnostics
```{r}
# Set layout for 2x2 grid of the 4 diagnostic plots
par(mfrow = c(2, 2)) 

# Create the 4 diagnostic plots for linear regression model
plot(lm_mortality_centered)

# Calculate Cook's Distance
cooksd <- cooks.distance(lm_mortality_centered)

# Plot Cook's Distance
plot(
  cooksd,
  main = "Cook's Distance for Mortality vs Deprivation × Screening",  # Plot title
  ylab = "Cook's Distance",  # Y-axis label
  type = "h",               # Use vertical lines for Cook's Distance
  col = "red",              # Make the lines red
  lwd = 2                   # Adjust the line width
)

# Add labels for each Local Authority
text(
  x = 1:length(cooksd), 
  y = cooksd, 
  labels = bowelcancer_screening_deprivation_mortality$`Local Authority`,  # Use Local Authority names
  cex = 0.7,                 # Adjust text size for readability
  pos = 4,                   # Position labels to the right of each point
  col = "blue"               # Set label colour to blue
)

# Add a threshold line for Cook's Distance (4/n rule)
threshold <- 4 / length(cooksd)  # Common threshold is 4/n
abline(h = threshold, col = "purple", lty = 2)  # Dashed purple line for threshold

```



```

